% Author: Anna (~0.5 page)
\section{Introduction}
Humans maintain an internal mental model of the world that allows them to simulate the consequences of events and actions without actually carrying them out.
This ability underlies core cognitive skills such as planning and decision-making.
In machine learning, world models learn the dynamics of an environment in order to predict future states from past observations, mapping a state at time $t$ to a state at time $t+1$ \parencite{https://doi.org/10.5281/zenodo.1207631}.
World Models therefore are equivalent to the internal models of humans \parencite{lecun2022path}.
In the classical formulation of Ha and Schmidhuber, such a model consists of three components: a vision component $V$ that compresses observations into a compact representation, a memory or dynamics component $M$ that predicts how this representation evolves over time, and a controller $C$ that selects actions based on the learned representation \parencite{https://doi.org/10.5281/zenodo.1207631}.
Classical approaches perform prediction directly in pixel space and condition the dynamics model on the actions of an agent.
More recent architectures such as the Joint Embedding Predictive Architecture (JEPA) instead perform prediction in a learned latent space, which avoids modelling irrelevant pixel level detail, and treat action conditioning as optional \parencite{lecun2022path}.
\\

As a testbed we use Lenia, a continuous generalisation of Conway's Game of Life \parencite{Chan2019Lenia, 9dca666c-8536-3d47-b0e4-6be2acc6bf4e}
Whereas classical cellular automata operate on binary cell states, Lenia uses continuous states in $[0,1]$ and continuous update rules, giving rise to rich, lifelike patterns \parencite{Chan2019Lenia}.
Lenia is a particularly attractive environment for studying world models as it contains no agent and hence no actions.
Its underlying physics is fully known making it possible to evaluate precisely whether a learned model has captured the true dynamics of the environment.
\\

This work addresses the following research question:
\begin{itemize}
    \item Does a world model trained in embedding space (JEPA style) learn better dynamics on Lenia than classical pixel space predictors?
\end{itemize}

Our work implements the vision and memory components of the world model framework.
Since Lenia contains no actions, no controller can be implemented.
This approach is consistent with the broader motivation behind world models \parencite{maes2026leworldmodelstableendtoendjointembedding}.
The objective is not to optimise behaviour for a specific task, but to learn representations that capture the dynamics of the environment.
Later on, the dynamics can be controlled or adapted to a diverse set of tasks.
Accordingly, we deliberately evaluate the vision and memory components only, building the task independent core of the world model.
By using controlled interventions as a proxy for the question of whether the learned model would be suitable for downstream planing.
Hence, building a controller is not part of this work.



